Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-07 paper_jasa.tex
worked "Appliation to Linear Models" over

\begin{lemma}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the classical test statistic. Suppose the associated test statistic in a classical test is observed as  $t_{obs}$ with $p$-value $p_D$.

@@

Note that the right hand side of equation (\ref{eqn:power_estimate}) is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

In this setting, there is a multitude of established graphics that help with the evaluation and diagnostics of a model \citep{cook:99}. Table \ref{tbl:stat_multiple} shows an overview of possible hypotheses that would be tested in the regression setting, and what visual test statistic might be used in these tests. These plots shoul be familiar, because they are all commonly used for evaluating and diagnosing regression models. For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would make a plot of residuals against fitted values (case 5).

@@

\begin{table*}[hbtp] 
\caption{Test Statistics for Testing Hypothesis Related to Model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for categorical $X_k$) & Box plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction with categorical $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.11}{\includegraphics{stat_interection.pdf}} \end{center} \end{minipage} & Scatter plot with least square lines of each category overlaid. For lineup plot, we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_nonlinear.pdf}} \end{center} \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For lineup plot, we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.15}{\includegraphics{stat_sigma_box.pdf}} \end{center} \end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For lineup plot, we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For lineup plot, we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_goodness_simple.pdf}} \end{center} \end{minipage} & Histogram of the response data. For lineup plot, we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid For lineup plot, we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

For the purposes of validating visual inference, we focus on cases 2 and 3, with a continuous and catgorical explanatory variable respectively.

Suppose $X_k$ is a categorical variable with two levels, and we test the hypothesis $H_0:\beta_k=0$ vs $H_1: \beta_k \ne 0$. If the responses for the two levels of the categorical variable $X_k$ in the model are significantly different and we fit the null model to the actual data, the resulting residual plot shows two groups of residuals. To test this we generate side-by-side boxplots of the residuals conditioned on the two levels of $X_k$, as the test statistic. If $\beta_k\ne 0$ the boxplots show a vertical displacement.

@@

Under the assumption that an observer is able to pick the plot with the smallest $p$-value from a lineup plot,  this leads to the decision to reject $H_0$ when $p_{B} < p_0$. Thus we have the expected power as

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{B} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}

The classical uniformly most powerful (UMP) test rejects $H_0$ if the observed test statistic $T= \hat{\beta_k}/ se(\hat{\beta_k})$ takes an extreme value in the distribution of theoretical $t$ distribution with $n-p$ degrees of freedom. Figure \ref{fig:power_expected} shows the power of the UMP test in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of UMP test. In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.65}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of UMP test for sample size $n =100$ and $\sigma = 12$. \hh{what is $m$? - would it make sense to show a curves for varying values of $m$?} \blue{$m$=20 here. When $m$ changes the type-I error also changes. Thus I can produce the difference between power curves for each $m$. Do you suggest that?}}
       \label{fig:power_expected}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=n+1, ...,n*c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$, $c$ is the proportion of contaminated data. For the experiment we consider $n=100$, $c=0.15$ producing actual data with $115$ points. Also $\alpha=0$ and $\lambda=10$, and $\sigma$ is chosen such that overall error standard deviation of the data is $5$ which is approximately 3.5\green{????} \blue{(Yes, this is how we did it. This value (3.5) makes the overall sd to be 5 because of the inclusion of the contaminated part of the data.)}. Figure \ref{fig:cont_dat} shows an example of contaminated data, generated by this model for $\beta=5$.  A linear model (Equation \eqref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment also follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null data plots are scatterplots of null data generated by simulating residuals from $N(0, {\hat{\sigma}}^2$) and plotting against $X_1$.

@@

\begin{figure}[hbt]
%\begin{figurehere}
   \centering
       \scalebox{0.4}{\includegraphics{contaminated_data.pdf}}
       \caption{Scatterplot of a simulated contaminated data set generated such that a simple linear model produce a slope almost negligible, ie, if we fit model (Equation \eqref{multi}) with $p=1$  to this data, the resulting estimate of $\beta_1$ would not be statistically significant.
}
       \label{fig:cont_dat}
\end{figure}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{power_observed.pdf}}
       \caption{Observed power of visual test from equation \eqref{power_est} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
       \label{fig:power_observed}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{power_observed_exp3.pdf}}
       \caption{Observed power of visual test from equation \eqref{power_est} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. }
       \label{fig:power_observed_exp3}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{p-value vs plot signal strength. The solid line represents 45 degree line which represents the equity shown in equation \eqref{plot_signal}}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.40}{\includegraphics{power_model.pdf}}
       \caption{Estimated power curve from equation \eqref{eqn:power} along with 95\% confidence interval for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
       \label{fig:power_model}
\end{figure}

>>>>

\begin{lemma}~\label{lemma}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the classical test statistic. Suppose the associated test statistic in a classical test is observed as  $t_{obs}$ with $p$-value $p_D$.

@@

Note that the right hand side of the equation (\ref{eqn:power_estimate}) is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and the visual test statistic that might be used for testing these hypotheses. These plots should be familiar, because they are all in common use. For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would make a plot of residuals against fitted values (case 5). For the purposes of comparing visual against classical inference, we focus on cases 2 and 3, with a continuous and catgorical explanatory variable respectively.

@@

\begin{table*}[hbtp] 
\caption{Visual test statistics for testing hypotheses related to the model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for categorical $X_k$) & Box plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction with categorical $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.11}{\includegraphics{stat_interection.pdf}} \end{center} \end{minipage} & Scatter plot with least square lines of each category overlaid. For lineup plot, we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_nonlinear.pdf}} \end{center} \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For lineup plot, we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.15}{\includegraphics{stat_sigma_box.pdf}} \end{center} \end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For lineup plot, we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}\begin{center}  	\scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For lineup plot, we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_goodness_simple.pdf}} \end{center} \end{minipage} & Histogram of the response data. For lineup plot, we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}\begin{center}  \scalebox{0.1}{\includegraphics{stat_intercept.pdf}} \end{center} \end{minipage} & Scatter plot with least square line overlaid For lineup plot, we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

Suppose $X_k$ is a categorical variable with two levels, and we test the hypothesis $H_0:\beta_k=0$ vs $H_1: \beta_k \ne 0$. If the responses for the two levels of the categorical variable $X_k$ in the model are different and we fit the null model to the actual data, the resulting residuals should show a significant difference between the two groups. To test this boxplots of the residuals conditioned on the two levels of $X_k$ are made, as the observed test statistic. If $\beta_k\ne 0$ the boxplots should show a vertical displacement.

@@

The classical test in this scenario uses $T= \hat{\beta_k}/
se(\hat{\beta_k})$ and rejects the null hypothesis if $T$ is extreme
on the scale of a $t$ distribution with $n-p$ degrees of freedom. This
is the uniformly most powerful (UMP) test \green{reference needed
  here} and it forms the benchmark upon which we evaluate the visual
test. To calculate what we might expect for the power of the visual
test, under perfect conditions, first assume that the observer is able
to pick the plot with the smallest $p$-value from a lineup plot.  This
leads to the decision to reject $H_0$ when $p_{B} < p_0$, where $p_{B}$ is calculated according to details given in Lemma \ref{lemma}. Thus 
the expected power of the visual test in this scenario is

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{B} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the UMP test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of UMP test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.65}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of UMP test for sample size $n =100$, $\sigma = 12$ and $m$=20. \green{Produce these curves for $m=12$, $m=20$, $m=30$.}}
       \label{fig:power_expected}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=n+1, ...,n*c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$, $c$ is the proportion of contaminated data. For the experiment we consider $n=100$, $c=0.15$ producing actual data with $115$ points. Also $\alpha=0$ and $\lambda=10$, and $\sigma$ is chosen such that overall error standard deviation of the data is $5$ which is approximately 3.5\green{????} \blue{(Yes, this is how we did it. This value (3.5) makes the overall sd to be 5 because of the inclusion of the contaminated part of the data.)}. Figure \ref{fig:cont_dat} shows an example of contaminated data, generated by this model for $\beta=5$.  A linear model (Equation \ref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment also follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null data plots are scatterplots of null data generated by simulating residuals from $N(0, {\hat{\sigma}}^2$) and plotting against $X_1$.

@@

\begin{figure}[hbt]
%\begin{figurehere}
   \centering
       \scalebox{0.4}{\includegraphics{contaminated_data.pdf}}
       \caption{Scatterplot of a simulated contaminated data set generated such that a simple linear model produce a slope almost negligible, ie, if we fit model (Equation \ref{multi}) with $p=1$  to this data, the resulting estimate of $\beta_1$ would not be statistically significant.
}
       \label{fig:cont_dat}
\end{figure}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{power_observed.pdf}}
       \caption{Observed power of visual test from equation \ref{power_est} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
       \label{fig:power_observed}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{power_observed_exp3.pdf}}
       \caption{Observed power of visual test from equation \ref{power_est} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. }
       \label{fig:power_observed_exp3}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{p-value vs plot signal strength. The solid line represents 45 degree line which represents the equity shown in equation \ref{plot_signal}}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.40}{\includegraphics{power_model.pdf}}
       \caption{Estimated power curve from Equation \ref{eqn:power} along with 95\% confidence interval for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
       \label{fig:power_model}
\end{figure}
